Model Recognition as Planning

### 2018-11-15 doc/icaps-19/model-recognition.tex
evaluation revised icaps19 paper

This section evaluates the empirical performance of {\em model recognition as planning}. To do so, we define a set $M$ of different planning models, that share the same state variables but that update these variables according to different action schemata. Then for each model $\mathcal{M}\in M$, we generate {\em partial observations} $\mathcal{O}(\tau)$ of {\em plan trajectories} that are computed with that model. Finally, we apply our {\em model recognition as planning} method to identify which model, among the models in $M$, was actually used for generating a given $\mathcal{O}(\tau)$ observation. For all the experiments, we assume equal prior probabilities, meaning $P(\mathcal{M}) = \frac{1}{|M|}$ for every $\mathcal{M} \in M$.

For the purpose of the experiments below, we introduce a particular class of $\mathcal{O}(\tau)$ observations. This new class allows us to distinguish between observable data, which may come from sensors, and hidden or latent factors that cannot be observed.
\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\mathcal{O}(\tau)$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

\subsubsection{Reproducibility.} All experiments were run in an Intel Core i5 3.10GHz x 4 16GB of RAM and the classical planner we used to solve the instances that result from the compilation was {\sc Madagascar}~\cite{rintanen2014madagascar} because of its ability to deal with classical planning problems with dead-ends~\cite{lopez2015deterministic}. We set up a timeout of 1000s for the computation of $\delta^o(\mathcal{M},\mathcal{O})$, at which point it is assigned maximum distance.

The compilation source code, evaluation scripts and benchmarks are fully available at this anonymous repository {\em } so any experimental data reported in the paper can be reproduced.

\subsection{Recognition of {\em Regular Automata}.} The first experiment, which doubles as a proof of concept, leverages {\em model recognition as planning} to perform a typical {\em string classification} problem. In this experiment, the system receives (1) the string to classify (the $\mathcal{O}(\tau)$ observation) and (2), a set $M$ of different planning models, where each $\mathcal{M}\in M$ codes a different class (i.e. a different regular automata that accepts any string that belong to the class).

Figure~\ref{fig:regautomatae} illustrates a three-symbol four-state {\em regular automata} for recognizing the $(abc)^+$ language. The {\em input alphabet} is $\Sigma=\{a,b,c\}$, and the machine states are $Q=\{q_0,q_1,q_2,\underline{q_3}\}$ (where \underline{$q_3$} is the only acceptor state). For instance, the execution of the planning model coding the {\em regular automata} defined in Figure~\ref{fig:regautomatae}, and with the input string $abcabc$, produces the following six-action plan {\small $(a,q_0\rightarrow q_1)$, $(b,q_1\rightarrow q_2)$, $(c,q_2\rightarrow q_3)$, $(a,q_3\rightarrow q_1)$, $(b,q_1\rightarrow q_2)$, $(c,q_2\rightarrow q_3)$}.

@@

In this experiment, $M$ comprises 5 planning models representing different regular automata defined with 5 states and a 4-symbol alphabet. The regular languages defined by these five automata are the following:
\begin{itemize}
	\item $\mathcal{L}$1: $a^+(b|c)d(dd)^*a$
	\item $\mathcal{L}$2: $bd(abd)^*cd^*c^+$
	\item $\mathcal{L}$3: $d^*c(ac)^*db^*ac^*b^+$
	\item $\mathcal{L}$4: $(cc)^+bd(abd)^*$
	\item $\mathcal{L}$5: $(d|a)a(ba)^*c^+d(dc^*d)^*$
\end{itemize}

For each regular language, we generated $20$ random strings (observations) of length 20 to 30  for a total of 100 strings. These observations contain no information about the structure of the possible automata. This means that $\mathcal{O}(\tau)$ is a $\Phi$-observation where fluents representing the internal machine state are unknown and the applied transitions (actions) are unobserved.

Table \ref{tab:conf_matrix} shows the confusion matrix resulting from classifying these 100 strings with {\em model recognition as planning}. In this matrix, rows represent the actual class and columns the class predicted by our {\em model recognition as planning} approach. Even though we are using a suboptimal classical planner, we can see that the class for all the 100 input strings was correctly recognized. These results proof that {\em model recognition as planning} is suitable for classification tasks where classes are given as generative planning models.

\begin{table}[hbt!]
	\centering
	\begin{tabular}{c|c|c|c|c|c}
		& $\mathcal{L}$1 & $\mathcal{L}$2 & $\mathcal{L}$3 & $\mathcal{L}$4 & $\mathcal{L}$5\\ \hline
		$\mathcal{L}$1 & 20 & 0 & 0 & 0 & 0 \\
		$\mathcal{L}$2 & 0 & 20 & 0 & 0 & 0 \\
		$\mathcal{L}$3 & 0 & 0 & 20 & 0 & 0 \\
		$\mathcal{L}$4 & 0 & 0 & 0 & 20 & 0 \\
		$\mathcal{L}$5 & 0 & 0 & 0 & 0 & 20 \\
	\end{tabular}
	\caption{Confusion matrix for {\em regular automata recognition}.}
	\label{tab:conf_matrix}
\end{table}

\subsection{Recognition of failures in {\em blocksworld} domain}
For our second experiment we have generated a set of observations for the {\em blocksworld} domain where some of the actions may fail. In more detail, the failures we consider are:

\begin{itemize}
	\item Stack fails and does nothing.
	\item Unstack fails and does nothing.
	\item Unstack fails and drops the block on the table.
\end{itemize}

We limited the failures to one type per observation, but it would be possible to recognize combinations of failures. The input models extend a 4-schemata {\em blocksworld} domain with an additional action schema that encodes one of the above failures. This means that the failure can be identified by finding the model that better explains the observation.

Our aim with this experiment is to validate the assumptions underlying our approximation of the $P(\mathcal{O}|\mathcal{M})$ likelihood. These assumption being that the agent acts rational, so the computation of the $P(\mathcal{O}|\mathcal{M})$ likelihood is dominated by the shortest trajectory that explains the observation, and that there exists a single such trajectory. In order to proof this, we assume that the length of the real trajectory is known, thus forcing the explanation to be optimal, and measure the accuracy of our approach across different degrees of observability. The accuracy metric is defined as the number of correct predictions over all predictions made, and the degree of observability indicates the probability of observing a literal in an intermediate state. Note that this definition allows for missing states (no literals were observed), but for this experiment we will make it so at least one literal was observed at every state so the length of the plan trajectory can be known from the observation. If our assumptions hold, we expect the results of this experiment to show high accuracy and to find an increasing trend as observability increases and the number of possible explanations is reduced.

%In this experiment, we measure the accuracy of our approach. Accuracy is defined as the number of correct predictions over all predictions made so, for this experiment, it translates to the percentage of correctly identified failures in the observations.

%We measure the accuracy across different degrees of observability, which indicate the probability of observing a literal in an intermediate state. Note that this definition allows for missing states (no literals were observed), but for this experiment we will make it so at least one literal was observed at every state so the length of the plan trajectory can be known from the observation, and in the next experiment we will forego this constraint. By knowing the length of the real trajectory we force the trajectory found by our approach to be rational, i.e, the shortest explanation for the observation. If our assumption holds, we expect the results to

\begin{figure}[hbt!]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/blocks.eps}
	\caption{Accuracy for the identification of failures in the \em{blocksworld} domain.}
	\label{fig:blocks_acc}
\end{figure}

Figure \ref{fig:blocks_acc} shows the accuracy of our approach in identifying failures over 30 observations. The figure showcases a positive trend proving that the more complete the observations are the more accurate our stimation becomes. The overall high accuracy also validates our assumption that the trajectory that better explains the observation is the shortest one.

\subsection{Recognition of navigation policies.} In this experiment the planning models given in $M$ represent different navigation models for $N\times N$ grids. In particular, we took a {\em classical navigation model} with actions {\tt\small up, down, right, left} (to move one cell in each of these four directions) and use it to define 8 different navigation policies with respect to these additional state variables {\tt\small \{(q0), (q1), (min ?v), (max ?v)\}}. An example is the navigation policy, shown in Figure~\ref{fig:model-example}. This policy allows to move right when $q_0$ holds while it allows to move left when $q_1$ holds, producing a zig-zag pattern when visiting all the cells of $N\times N$ grids.

On the other hand we generated observations from 8 different trajectories, one for each of the 8 previous planning models. The trajectories depict the path followed by each navigation policy to solve the planning problem of {\em visiting all} cells in a $5\times 5$ grid. As in the previous experiments, observations contain no actions and here, for each observed state, only the values of the fluents encoding the {\em x and y coordinates} of the agent are known (i.e. $\Phi =$\{{\small\tt xcoord, ycoord}\}).

Again, this experiment evaluates the accuracy of {\em model recognition as planning} to identify the model that generated each observation under different degrees of observability. The novel aspect of this experiment is that now, we do not guarantee that observed states contain at least one literal, meaning that some intermediate states may be missing. Therefore, in this new scenario more than a single action may be required to reach one observed state from another.

Another interesting aspect of this experiment is that all navigation policies are at a maximum distance of just 4 editions from the base {\em classical navigation model} that can move in any direction at any given state. This means that, for any given model and observation, we will need to perform a maximum of 4 editions to the input model to explain the observation. This aspect heavily constrains the discriminating power of the $P(\mathcal{O}|\mathcal{M})$ likelihood, and added to the low observability of the states produces a very benchmark for {\em model recognition as planning}.

@@

\begin{figure}[hbt!]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/navigation.eps}
	\caption{Classification accuracy for the recognition of navigation models.}
	\label{fig:navigation_acc}
\end{figure}

@@

Previous work on the learning of \strips\ action models also defined semantic error metrics to quantify the errors of a model with respect to the observation of a plan execution~\cite{yang2007learning}. Our approach for quantifying this error is based on the definition of a edit distance for the model which allow us to not accumulate the repetition of errors coming from the the same model flaw. A related approach is recently followed for {\em model reconciliation}~\cite{ChakrabortiSZK17} where model edition is used to conform the PDDL models of two different agents. Also  related to this paper is the work on {\em Goal Recognition Design} but in that work the action model is given in advance and fixed~\cite{KerenGK14}.

>>>>

This section presents three different applications of {\em model recognition} and evaluates the empirical performance of {\em model recognition as planning} when addressing the three of them.

In all the experiments we define a set $M$ of different planning models, that share the same state variables but that update these variables according to different action schemata. Then for each model $\mathcal{M}\in M$, we generate {\em partial observations} $\mathcal{O}(\tau)$ of {\em plan trajectories} that were computed with that model. Finally, we apply our {\em model recognition as planning} method to identify which model, among the models in $M$, was actually used for generating a given $\mathcal{O}(\tau)$ observation. For all the experiments, we assume equal prior probabilities, meaning $P(\mathcal{M}) = \frac{1}{|M|}$ for every $\mathcal{M} \in M$.

To provide a better explanation of the experiments below, we introduce a particular class of $\mathcal{O}(\tau)$ observations. This new class allows us to distinguish between {\em observable} state variables, whose value may come from sensors, and {\em hidden} or {\em latent} state variables that cannot be observed.
\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\mathcal{O}(\tau)$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}
Therefore the value of variables in $\Phi\subseteq F$ can be {\em observed} while the value of the variables in $F\setminus\Phi$ is {\em hidden}.

\subsubsection{Reproducibility.} All experiments were run in an Intel Core i5 3.10GHz x 4 16GB of RAM and the classical planner we used to solve the instances that result from the compilation was {\sc Madagascar}~\cite{rintanen2014madagascar} because of its ability to deal with classical planning problems with dead-ends~\cite{lopez2015deterministic}. We set up a timeout of 1000s for the computation of $\delta^o(\mathcal{M},\mathcal{O})$, at which point it is assigned maximum distance. Compilation source code, evaluation scripts and benchmarks are fully available at this anonymous repository {\em } so any experimental data reported in the paper can be reproduced.

\subsection{Recognition of {\em regular automata}.} The first experiment, which doubles as a proof of concept, leverages {\em model recognition as planning} to perform a typical {\em string classification} problem. In this experiment, the system receives (1) the string to classify (the $\mathcal{O}(\tau)$ observation) and (2), a set $M$ of different planning models, where each $\mathcal{M}\in M$ codes a different class (i.e. a different regular automata that accepts any string that belong to the class).

Figure~\ref{fig:regautomatae} illustrates a three-symbol four-state {\em regular automata} for recognizing the $(abc)^+$ language. The {\em input alphabet} is $\Sigma=\{a,b,c\}$, and the machine states are $Q=\{q_0,q_1,q_2,\underline{q_3}\}$ (where \underline{$q_3$} is the only acceptor state). For instance, the execution of the planning model coding the {\em regular automata} defined in Figure~\ref{fig:regautomatae}, and with the input string $abcabc$, produces the following six-action plan {\small $(\langle a,q_0\rangle\rightarrow q_1)$, $(\langle b,q_1\rangle\rightarrow q_2)$, $(\langle c,q_2\rangle\rightarrow \underline{q_3})$, $(\langle a,\underline{q_3}\rangle\rightarrow q_1)$, $(\langle b,q_1\rangle\rightarrow q_2)$, $(\langle c,q_2\rangle\rightarrow \underline{q_3})$}.

@@

In this experiment, $M$ comprises 5 planning models representing different regular automata defined with 5 states and a 4-symbol alphabet. The five regular languages defined by these five automata are the following:
\begin{itemize}
	\item $\mathcal{L}$1: $a^+(b|c)d(dd)^*a$
	\item $\mathcal{L}$2: $bd(abd)^*cd^*c^+$
	\item $\mathcal{L}$3: $d^*c(ac)^*db^*ac^*b^+$
	\item $\mathcal{L}$4: $(cc)^+bd(abd)^*$
	\item $\mathcal{L}$5: $(d|a)a(ba)^*c^+d(dc^*d)^*$
\end{itemize}

For each regular language, we generated $20$ random strings, $\mathcal{O}(\tau)$ observations, of length 20 to 30 for a total of 100 strings. More precisely observations $\mathcal{O}(\tau)$ are $\Phi$-observations where: the applied automata transitions and the fluents representing the internal automata state are unobserved. Table \ref{tab:conf_matrix} shows the {\em confusion matrix} resulting from classifying these 100 strings with {\em model recognition as planning}. In this matrix, rows represent the actual class and columns the class predicted by our {\em model recognition as planning} approach. Even though we are using a suboptimal classical planner, we can see that the class for all the 100 input strings was correctly recognized. These results proof the suitability of {\em model recognition as planning} for classification tasks where classes are given as generative planning models and examples are sequence of observations.

\begin{table}
	\centering
	\begin{tabular}{c|c|c|c|c|c}
		& $\mathcal{L}$1 & $\mathcal{L}$2 & $\mathcal{L}$3 & $\mathcal{L}$4 & $\mathcal{L}$5\\ \hline
		$\mathcal{L}$1 & 20 & 0 & 0 & 0 & 0 \\
		$\mathcal{L}$2 & 0 & 20 & 0 & 0 & 0 \\
		$\mathcal{L}$3 & 0 & 0 & 20 & 0 & 0 \\
		$\mathcal{L}$4 & 0 & 0 & 0 & 20 & 0 \\
		$\mathcal{L}$5 & 0 & 0 & 0 & 0 & 20 \\
	\end{tabular}
	\caption{Confusion matrix for {\em regular automata recognition}.}
	\label{tab:conf_matrix}
\end{table}

\subsection{Recognizing failures in a non-deterministic {\em blocksworld}}
With this second experiment we aim to validate the assumptions underlying our approximation of the $P(\mathcal{O}|\mathcal{M})$ likelihood. These assumptions being that the agent acts rational, so $P(\mathcal{O}|\mathcal{M})$ is dominated by the shortest trajectory that explains the observation, and that there exists a single such trajectory.

In this particular experiment we assume that the length of the trajectory is known (to force the explanation to be optimal) and hence, measure the {\em  accuracy} of our approximation of the $P(\mathcal{O}|\mathcal{M})$ likelihood across different degrees of observability. More precisely, we assume that at least one literal is observed at every state so the length of the plan trajectory is fixed by the observation.

The input observations $\mathcal{O}(\tau)$ come from a non-deterministic {\em blocksworld} in which some of the executed actions failed. We considered failures of three kinds:
\begin{itemize}
	\item $\mathcal{F}$1: the execution of a {\tt\small stack} action fails causing no effect.
	\item $\mathcal{F}$2: {\tt\small unstack} fails dropping the block to unstack on the table.        
	\item $\mathcal{F}$3: {\tt\small unstack} fails doing nothing.
\end{itemize}
The input models in $M$ are then three different {\em extended} 4-operator {\em blocksworlds} such that each of these models includes an additional action schema that encodes one of the three failures above. This means that failures are identified by finding the model that explains the best the input observation.

Figure \ref{fig:blocks_acc} shows the {\em accuracy} of our approach when identifying failures over 30 different observations. {\em Accuracy} is defined as the number of correct predictions over all predictions made while the {\em degree of observability} indicates the probability of observing a literal in an intermediate state. The figure showcases a positive trend proving that the more complete the observations are the more accurate our estimation becomes. The overall high accuracy validates our assumption that the trajectory that better explains the observation is the shortest one. %We limited the failures to be of one kind per observation however, it would be possible to recognize different combinations of them.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{figures/blocks.eps}
	\caption{Accuracy for the recognition of failures in a non-deterministic {\em blocksworld}.}
	\label{fig:blocks_acc}
\end{figure}

\subsection{Recognition of navigation policies}
Again, this experiment evaluates the accuracy of {\em model recognition as planning} to identify the model that generated each observation under different {\em degrees of observability}. The novel aspect of this experiment is that now, we do not guarantee that observed states contain at least one literal, meaning that some intermediate states may be missing. Therefore, in this new scenario more than a single action may be required to reach one observed state from another.

In this experiment the planning models given in $M$ represent different navigation models for $N\times N$ grids. In particular, we took a {\em classical navigation model} with actions {\tt\small up, down, right, left} (to move one cell in each of these four directions) and extend it to define 8 different navigation policies with respect to these additional state variables {\tt\small \{(q0), (q1), (min ?v), (max ?v)\}}. An example is the navigation policy, shown in Figure~\ref{fig:model-example}. This policy allows to move right when $q_0$ holds while it allows to move left when $q_1$ holds, producing a zig-zag pattern when visiting all the cells of $N\times N$ grids.

Note that all navigation policies are at a maximum distance of just 4 editions from the base {\em classical navigation model} that can move in any direction at any given state. This aspect constrains the discriminating power of the $P(\mathcal{O}|\mathcal{M})$ likelihood, and added to the low observability of the states, produces a touchstone for {\em model recognition as planning}.

On the other hand we generated observations from 8 different trajectories, one for each of the 8 previous planning models. The trajectories depict paths followed by each navigation policy to solve the planning problem of {\em visiting all} cells in a $5\times 5$ grid. As in the previous experiments, observations contain no actions and here, for each observed state, only the values of the fluents encoding the {\em x and y coordinates} of the agent are known (i.e. $\Phi =$\{{\small\tt xcoord, ycoord}\}).

@@

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{figures/navigation.eps}
	\caption{Classification accuracy for the recognition of navigation models.}
	\label{fig:navigation_acc}
\end{figure}

\section{Related work}
\label{sec:related}
A related approach is recently followed for {\em model reconciliation}~\cite{ChakrabortiSZK17} where model edition is used to conform the PDDL models of two different agents. Also  related to this paper is the work on {\em Goal Recognition Design} but in that work the action model is given in advance and fixed~\cite{KerenGK14}.

Previous work on the learning of \strips\ action models also defined semantic error metrics to quantify the errors of a model with respect to the observation of a plan execution~\cite{yang2007learning}. Our approach for quantifying this error is based on the definition of a edit distance for the model which allow us to not accumulate the repetition of errors coming from the the same model flaw.

@@
